Brief Announcement: An Improved Lower Bound for Dynamic Balanced Graph Partitioning

### 2020-02-14 paper/paper.tex
Better notation

\begin{abstract}
  In the online Graph RePartitioning problem,
  a sequence of communication requests arrives online,
  between pairs of $n$ nodes that are partitioned into $\ell$ clusters
  each of size $k$ (i.e., $n = \ell \cdot k$).
  The objective is to serve these requests while paying a cost that is reasonably close to the optimal cost of serving them given  the whole sequence in advance. 
  This is achieved by opting a sequence of  partitions (a.k.a.~configurations)
  such that frequently communicating nodes (from different clusters)
  are relocated to the same cluster before they inflict too much communication cost.
  A \emph{reconfiguration} (i.e., re-partitioning) is performed by
  moving (i.e., migrating) nodes between clusters at a cost.
  The goal is to jointly minimize communication and migration costs while serving the sequence.

	The best known deterministic algorithm is $O(k^2\cdot \ell^2)$-competitive \cite{?????},
	and the best known lower bound for the competitive ratio of
	deterministic algorithms is $\Omega(k)$,
	which uses a reduction from the online paging problem.
	  In this paper, we provide the lower bound of $\Omega(k\cdot \ell)$.
    We also present an optimal algorithm for the case of $k=3$.
  Finally,
  we provide a $O(k\cdot \ell)$-competitive algorithm for the variant
  where the optimal configuration is always a fixed partition
  that incurs zero communication cost through the whole request sequence.
\end{abstract}

@@

Formally, the online \emph{Balanced RePartitioning} problem (\OBRP{}) is defined as
follows. We are given a set of $n$ nodes,
initially arbitrarily partitioned into $\ell$~clusters,
each of size~$k$.
We say that nodes in $C \subset V$ are \emph{collocated}
% in a partition $P$
if they are assigned to the same cluster.
%\maciek{We can ommit the formal definition: "I.e.,
%$\forall u,v \in C: \mathit{cluster} (u,P) = \mathit{cluster} (v,P)$.". %However, if we use the notion of cluster later, then TODO: define cluster}
%\maciek{We can ommit the formal definition: "I.e.,
%$\forall u,v \in C: \mathit{cluster} (u,P) = \mathit{cluster} (v,P)$.". However, if we use the notion of cluster later, then TODO: define cluster}
The input consists of a sequence of pairwise communication requests
$\sigma = (u_1,v_1),$ $(u_2,v_2),$ $(u_3,v_3), \ldots$,
where a pair $(u_t,v_t)$ indicates that nodes $u_t$ and $v_t$ exchange a fixed amount of data.
Before serving the request,
the online algorithm can re-partition,
i.e.,
it may (optionally) move some of the nodes into clusters different than their current assigned clusters.
Once the re-partition is executed,
we serve
the communication request between two  nodes at cost~0 if they are collocated,
or at cost~1 if they are located in different clusters.
The cost of migrating a node from one cluster to another
is~$\alpha \in \mathbb{Z}^+$.
For any algorithm $ALG$,
its cost,
denoted by $ALG(\sigma)$,
is the total cost of communications and
the cost of migrations performed by $ALG$ while serving the sequence $\sigma$.

@@

  \section{An Optimal Algorithm for $k=3$ (most likely wrong)}

@@

  \begin{theorem}
    \TAlg{} is $O(\ell)$-competitive.
  \end{theorem}
  \begin{proof}
    Fix a phase of \TAlg, and consider the state of its counters at the end of it.

    Let $A^I$ be the cost of (extra-cluster) communication incurred in this phase by \TAlg between pairs that belonged to the single component at the end of the phase.
    Let $A^E$ be the cost of (extra-cluster) communication incurred in this phase between the nodes that belong to different components throughout this phase.
    Let $A^M$ be the cost of migrations performed by \TAlg in this phase.

    We say that the pair $(u, v)$ is \emph{saturated}, if the counter has value $\alpha$, and \emph{unsaturated} otherwise.

    During a single phase, at most $2\cdot \ell$ component merge operations are performed.
    Exceeding this number would mean that a component of size $4$ would emerge, a contradiction.
    Together with Lemma~\ref{lem:1req}, this gives us $A^M \leq 6\cdot\alpha\cdot\ell$.

    Now we bound $A^I$.
    Each component of size $3$ contributes at most $(3\cdot \alpha - 1)$ to $A^I$, as at most $2$ of its pairs are saturated and contribute $\alpha$ each, and the unsaturated pair contributes at most $\alpha-1$.
    Each of $\ell$ clusters of \TAlg contain at most one component of size $3$ (other cluster configurations contribute less: $3$ singletons contribute $0$ and a component of size $2$ contributes at most $\alpha$ to $A^I$), and in total, $A^I \leq \ell \cdot (3\cdot \alpha-1)$.

    We bound $A^E$ by $k^2 \cdot (\alpha - 1)$, as no more than $k^2$ pairs are unsaturated, and each of them contributes at most $\alpha -1$.

    In total, the cost of \TAlg is at most $A^I + A^E + A^M$ during the phase.

    Now we lower bound the cost of $\OPT$.
  \end{proof}

>>>>

\begin{abstract}
  In the online Graph RePartitioning problem, we are given a sequence of pairwise communication requests between n nodes, revealed online.
  The objective is to service these requests by partitioning the nodes into $\ell$ clusters, each of size $k$, such that frequently communicating nodes are located in the same cluster.
  The partitioning can be updated by migrating nodes between clusters for a fixed cost.
  The goal is to jointly minimize the amount of inter-cluster communication and migration cost.

  So far, for this natural problem, the best known deterministic algorithm was $O(k^2\cdot \ell^2)$-competitive, and the best known lower bound was $\Omega(k)$ for the competitive ratio of any deterministic algorithm that used the reduction from the caching problem.
  In this paper, we provide the lower bound of $\Omega(k\cdot \ell)$.

  We also present the optimal algorithm for $k=3$.
  Finally, we provide a $O(k\cdot \ell)$-competitive algorithm for the scenario, where the perfect partition of input exists.
\end{abstract}

@@

Formally, the online \emph{Balanced RePartitioning} problem (\OBRP{}) is defined as
follows. We are given a set of $n$ nodes,
initially arbitrarily partitioned into $\ell$~clusters,
each of size~$k$.
We say that nodes in $C \subset V$ are \emph{collocated}% in a partition $P$
if they are assigned to the same cluster.
%\maciek{We can ommit the formal definition: "I.e.,
%$\forall u,v \in C: \mathit{cluster} (u,P) = \mathit{cluster} (v,P)$.". %However, if we use the notion of cluster later, then TODO: define cluster}
%\maciek{We can ommit the formal definition: "I.e.,
%$\forall u,v \in C: \mathit{cluster} (u,P) = \mathit{cluster} (v,P)$.". However, if we use the notion of cluster later, then TODO: define cluster}
The input consists of a sequence of pairwise communication requests
$\sigma = (u_1,v_1),$ $(u_2,v_2),$ $(u_3,v_3), \ldots$,
where a pair $(u_t,v_t)$ indicates that nodes $u_t$ and $v_t$ exchange a fixed amount of data.
Before serving the request,
the online algorithm can re-partition,
i.e.,
it may (optionally) move some of the nodes into clusters different than their current assigned clusters.
Once the re-partition is executed,
we serve
the communication request between two  nodes at cost~0 if they are collocated,
or at cost~1 if they are located in different clusters.
The cost of migrating a node from one cluster to another
is~$\alpha \in \mathbb{Z}^+$.
For any algorithm $ALG$,
its cost,
denoted by $ALG(\sigma)$,
is the total cost of communications and
the cost of migrations performed by $ALG$ while serving the sequence $\sigma$.

@@

  \section{An Optimal Algorithm for $k=3$}

@@

  \begin{theorem}
    \TAlg{} is $O(\ell)$-competitive.
  \end{theorem}
  \begin{proof}
    Fix a completed phase of \TAlg, and consider the state of its counters at the end of it.
    We say that the pair $(u, v)$ is \emph{saturated}, if the counter has value $\alpha$, and \emph{unsaturated} otherwise.

    We partition the input sequence $\sigma$ that arrived during the phase into subsequences $\sigma_I$ (called \emph{internal}) and $\sigma_E$ (called \emph{external}), defined as follows.
    Requests $\sigma_E$ concern pairs of nodes that belong to the same component of \TAlg at the end of the phase, and $\sigma_E$ denotes all other requests issued during this phase.

    %Let $A^I$ be the cost of (extra-cluster) communication incurred in this phase by \TAlg between pairs that belonged to the single component at the end of the phase.
    %Let $A^E$ be the cost of (extra-cluster) communication incurred in this phase between the nodes that belong to different components at the end of this phase.
    Let $ALG(M)$ be the cost of migrations performed by \TAlg in this phase.

    During a single phase, at most $2\cdot \ell$ component merge operations are performed.
    Exceeding this number would mean that a component of size $4$ would emerge, a contradiction.
    Together with Lemma~\ref{lem:1req}, this gives us $ALG(M) \leq 6\cdot\alpha\cdot\ell$.

    Now we bound $ALG(\sigma_I)$.
    Each component of size $3$ contributes at most $(3\cdot \alpha - 1)$ to $A^I$, as at most $2$ of its pairs are saturated and contribute $\alpha$ each, and the unsaturated pair contributes at most $\alpha-1$.
    Each of $\ell$ clusters of \TAlg contain at most one component of size $3$ (other cluster configurations contribute less: $3$ singletons contribute $0$ and a component of size $2$ contributes at most $\alpha$ to $A^I$), and in total, $ALG(\sigma_I) \leq \ell \cdot (3\cdot \alpha-1)$.

    %We bound $A^E$ by $k^2 \cdot (\alpha - 1)$, as no more than $k^2$ pairs are unsaturated, and each of them contributes at most $\alpha -1$.
    %\maciek{not needed most likely}

    In total, the cost of \TAlg is at most $ALG(\sigma_I) + ALG(\sigma_E) + ALG(M) \leq 9\alpha\cdot \ell + ALG(\sigma_E)$ during this phase.

    \medskip

    Now we lower bound the cost of $\OPT$ by $\OPT(\sigma_I) +\OPT(\sigma_E)+\OPT(M)$, defined with respect to components of \TAlg.
    We split the cost of $\OPT$ into parts coming from serving $\sigma_I$ and $\sigma_E$.
    While serving these requests, $\OPT$ may perform migrations, and we account for them in both parts.

    %Let $O^M$ be the cost of migrations performed by $\OPT$ during the phase.
    %Let $O^I$ be the cost of serving requests between nodes that were put in one component by \TAlg during this phase.
    %Let $O^E$ be the cost serving requests between nodes that \TAlg did not put in the same component during that phase.

    We have $OPT(M) + OPT(\sigma_I) \geq \alpha$, as the components of \TAlg could not be partitioned without splitting them, and there exist a pair of nodes with at least $\alpha$ requests between them.
    $\OPT$ either served them remotely or performed a migration.

    Now we bound $\OPT(M) + \OPT(\sigma_E))$.

    \medskip

    Now we consider the last, unfinished phase.
    If the unfinished phase is also the first one, then we use the fact that we started with the same initial configuration.
    Then, we charge $\OPT$ $\alpha$ for first $\alpha$ requests or a migration,
    and we follow the analysis regarding the unsaturated requests.
    \maciek{todo}

    If we have at least two phases, then we assume that OPT did not pay during that, and the cost of ALG increases at most twofold.
  \end{proof}
